---
title: "Working with data.frames"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
description: "An introduction to data.frames -- indexing (subsetting) tabular data in base R."

---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```


## What is a data frame?

To this point we have worked with vectors of data.  data.frames in R allow us to work with tabular data -- that is, data arranged in row and columns.  But really, under the hood, data.frames are just collections of equal-length vectors: each column in a data.frame is a vector.

Here is the `iris` data.frame.  It's built in to R, so it's always available just by typing `iris`

```{r, echo=T}
iris
```

In the console, type `?iris` to learn more about the `iris` data.frame.

###

In lecture we learned that you can extract columns from a data.frame using in two different ways -- using `$` and `[[`

- extract the `Species` column from the `iris` data.frame using `$`

### Exercise 

*Here's a simple exercise with an empty code chunk provided for entering the answer.*

Write the R code required to add two plus two:

```{r two-plus-two, exercise=TRUE}

```

### Exercise with Code

*Here's an exercise with some prepopulated code as well as `exercise.lines = 5` to provide a bit more initial room to work.*

Now write a function that adds any two numbers and then call it:

```{r add-function, exercise=TRUE, exercise.lines = 5}
add <- function() {
  
}
```

## Topic 2

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaluated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```
